---
title: "Comparing CSDID Estimators: Algorithms, Base Periods, and Software"
author: ""
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{geometry}
  - \geometry{margin=2.5cm}
  - \captionsetup[table]{skip=5pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "95%"
)
```

```{r packages}
# ── Core DiD packages ──────────────────────────────────────────────────────────
library(did)         # Callaway & Sant'Anna (C&S) — att_gt()
library(fixest)      # Sun & Abraham via sunab()
library(staggered)   # Roth & Sant'Anna staggered estimator

# ── Output / helpers ──────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(patchwork)

# ── Stata integration (requires Stata on PATH) ────────────────────────────────
# library(Statamarkdown)   # uncomment if Stata is available
```

---

# Data

We use the `mpdta` dataset from the **did** package — a balanced panel of
county-level teen employment rates across 500 counties, where counties are
treated at different calendar years (staggered adoption).

```{r data}
data(mpdta, package = "did")

# Quick overview
head(mpdta)
cat(sprintf(
  "Counties: %d | Years: %s | Treatment cohorts: %s\n",
  n_distinct(mpdta$countyreal),
  paste(sort(unique(mpdta$year)),    collapse = ", "),
  paste(sort(unique(mpdta$first.treat[mpdta$first.treat > 0])), collapse = ", ")
))
```

The outcome is `lemp` (log teen employment), the unit ID is `countyreal`,
the time variable is `year`, and `first.treat` gives the first treatment period
(0 = never treated).

---

# Part A: Comparing Estimation Algorithms (`did::att_gt`)

`att_gt()` from the **did** package supports three estimation methods:

| Method | Description |
|--------|-------------|
| `"dr"` | Doubly-Robust IPW (default) — combines outcome regression + propensity score |
| `"ipw"` | Inverse Probability Weighting only |
| `"reg"` | Outcome Regression only |

```{r algo-est, cache=TRUE}
run_attgt <- function(method, control = "nevertreated") {
  did::att_gt(
    yname         = "lemp",
    tname         = "year",
    idname        = "countyreal",
    gname         = "first.treat",
    data          = mpdta,
    est_method    = method,
    control_group = control
  )
}

fit_dr   <- run_attgt("dr")
fit_ipw  <- run_attgt("ipw")
fit_reg  <- run_attgt("reg")
```

## ATT(g,t) Comparison

```{r algo-tbl, results='asis'}
extract_att <- function(fit, label) {
  ag <- did::aggte(fit, type = "simple")
  data.frame(
    Method   = label,
    ATT      = round(ag$overall.att, 4),
    SE       = round(ag$overall.se,  4),
    CI_lower = round(ag$overall.att - 1.96 * ag$overall.se, 4),
    CI_upper = round(ag$overall.att + 1.96 * ag$overall.se, 4)
  )
}

algo_tbl <- bind_rows(
  extract_att(fit_dr,  "DR-IPW"),
  extract_att(fit_ipw, "IPW"),
  extract_att(fit_reg, "Regression")
)

kable(
  algo_tbl,
  caption = "Overall ATT by Estimation Algorithm (csdid, never-treated control)",
  booktabs = TRUE, linesep = ""
) |>
  kable_styling(latex_options = c("hold_position"))
```

## Event-Study Plots by Algorithm

```{r algo-event, fig.height=7}
es_plot <- function(fit, title) {
  es <- did::aggte(fit, type = "dynamic")
  df <- data.frame(
    e    = es$egt,
    att  = es$att.egt,
    se   = es$se.egt
  ) |>
    mutate(lo = att - 1.96*se, hi = att + 1.96*se)

  ggplot(df, aes(x = e, y = att)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = -0.5, linetype = "dotted", colour = "grey50") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
    geom_line(colour = "steelblue") +
    geom_point(colour = "steelblue") +
    labs(title = title, x = "Event time", y = "ATT") +
    theme_bw(base_size = 11)
}

p1 <- es_plot(fit_dr,  "DR-IPW")
p2 <- es_plot(fit_ipw, "IPW")
p3 <- es_plot(fit_reg, "Regression")

(p1 | p2 | p3) +
  plot_annotation(title = "Event-Study: Algorithm Comparison",
                  theme = theme(plot.title = element_text(size = 13, face = "bold")))
```

**Takeaway.** The DR-IPW estimator is doubly robust — it remains consistent if
either the propensity score or the outcome model is correctly specified. The
three estimators typically agree closely on well-behaved data; divergence signals
model mis-specification.

---

# Part B: Comparing Base Periods (Control Group)

C&S allows two choices for the comparison group:

| Option | Description |
|--------|-------------|
| `"nevertreated"` | Only units never treated serve as control |
| `"notyettreated"` | Units not yet treated at time $t$ serve as control |

The choice matters when the "not-yet-treated" group may itself be partially
contaminated by anticipation or when never-treated units are few.

```{r base-est, cache=TRUE}
fit_never  <- run_attgt("dr", control = "nevertreated")
fit_notyet <- run_attgt("dr", control = "notyettreated")
```

## Overall ATT Comparison

```{r base-tbl, results='asis'}
base_tbl <- bind_rows(
  extract_att(fit_never,  "Never treated"),
  extract_att(fit_notyet, "Not yet treated")
)

kable(
  base_tbl,
  caption = "Overall ATT by Control Group Choice (DR-IPW)",
  booktabs = TRUE, linesep = ""
) |>
  kable_styling(latex_options = c("hold_position"))
```

## Event-Study Plots by Base Period

```{r base-event, fig.height=4.5}
p_never  <- es_plot(fit_never,  "Never treated")
p_notyet <- es_plot(fit_notyet, "Not yet treated")

(p_never | p_notyet) +
  plot_annotation(title = "Event-Study: Control Group Comparison",
                  theme = theme(plot.title = element_text(size = 13, face = "bold")))
```

## ATT(g,t) Heat Map

```{r base-heat, fig.height=5}
att_matrix <- function(fit, label) {
  data.frame(
    g      = fit$group,
    t      = fit$t,
    att    = fit$att,
    source = label
  )
}

heat_df <- bind_rows(
  att_matrix(fit_never,  "Never treated"),
  att_matrix(fit_notyet, "Not yet treated")
) |>
  filter(g > 0)

ggplot(heat_df, aes(x = factor(t), y = factor(g), fill = att)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "steelblue",
                       midpoint = 0, name = "ATT(g,t)") +
  facet_wrap(~source) +
  labs(title = "ATT(g,t) by Control Group",
       x = "Calendar time t", y = "Cohort g") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Takeaway.** With many treated cohorts and few never-treated units,
`"notyettreated"` gives more precise estimates but requires assuming no
anticipation among later-treated units. The two choices converge when the panel
is large and staggering is moderate.

---

# Part C: Comparing Software Packages

## R Packages

We compare three R packages against the baseline `did::att_gt()` DR result
already estimated in Part A (`fit_dr`):

| Package | Estimator | Notes |
|---------|-----------|-------|
| `did` (`att_gt`) | Callaway & Sant'Anna (2021) | DR-IPW, never-treated; baseline |
| `fixest` | Sun & Abraham (2021) via `sunab()` | TWFE-based, fast |
| `staggered` | Roth & Sant'Anna (2023) | Efficiency-optimised |

### `did` (baseline — already estimated)

```{r pkg-did}
att_did <- did::aggte(fit_dr, type = "simple")
cat(sprintf("did::att_gt | ATT = %.4f | SE = %.4f\n",
            att_did$overall.att, att_did$overall.se))
```

### `fixest` (Sun & Abraham)

```{r pkg-fixest}
# Encode cohort for sunab()
mpdta2 <- mpdta |>
  mutate(cohort = if_else(first.treat == 0, Inf, as.numeric(first.treat)))

fit_sa <- feols(
  lemp ~ sunab(cohort, year) | countyreal + year,
  data    = mpdta2,
  cluster = ~countyreal
)
# Aggregate: simple average over all (cohort, t) post-treatment cells
coefs_sa <- coef(fit_sa)
se_sa    <- se(fit_sa)
post     <- as.numeric(sub(".*::([-0-9]+)$", "\\1", names(coefs_sa)))
post_idx <- post >= 0

att_sa <- mean(coefs_sa[post_idx])
se_sa_agg <- sqrt(mean(se_sa[post_idx]^2))   # conservative; for SE use iplot()
cat(sprintf("fixest  | ATT = %.4f | SE ≈ %.4f (naive avg)\n", att_sa, se_sa_agg))
```

### `staggered`

```{r pkg-staggered, cache=TRUE}
fit_stag <- staggered::staggered(
  df       = mpdta,
  i        = "countyreal",
  t        = "year",
  g        = "first.treat",
  y        = "lemp",
  estimand = "simple"
)
cat(sprintf("staggered | ATT = %.4f | SE = %.4f\n",
            fit_stag$estimate, fit_stag$se))
```

### Summary Table

```{r pkg-summary, results='asis'}
pkg_tbl <- tribble(
  ~Package,     ~Estimator,          ~ATT,                 ~SE,
  "did",        "C&S DR-IPW",        att_did$overall.att,  att_did$overall.se,
  "fixest",     "Sun & Abraham",     att_sa,               se_sa_agg,
  "staggered",  "Roth & Sant'Anna",  fit_stag$estimate,    fit_stag$se
) |>
  mutate(
    ATT      = round(ATT, 4),
    SE       = round(SE, 4),
    CI_lower = round(ATT - 1.96*SE, 4),
    CI_upper = round(ATT + 1.96*SE, 4)
  )

kable(
  pkg_tbl,
  caption = "Overall ATT by R Package",
  booktabs = TRUE, linesep = ""
) |>
  kable_styling(latex_options = c("hold_position"))
```

### Coefficient Plot

```{r pkg-coefplot, fig.height=3.5}
ggplot(pkg_tbl, aes(x = Package, y = ATT,
                    ymin = CI_lower, ymax = CI_upper,
                    colour = Package)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(size = 0.8, linewidth = 0.8) +
  coord_flip() +
  scale_colour_brewer(palette = "Set1", guide = "none") +
  labs(title = "Overall ATT: R Package Comparison",
       x = NULL, y = "ATT (95% CI)") +
  theme_bw(base_size = 11)
```

---

## Stata via `Statamarkdown`

The chunk below runs **Stata** code directly inside R Markdown using the
`Statamarkdown` package. Requires Stata to be installed and its executable on
`PATH`. Set `eval=TRUE` when running on a machine with Stata.

```{stata stata-setup, eval=FALSE, collectcode=TRUE}
* ── Install csdid if needed ──────────────────────────────────────────────────
* ssc install csdid, replace
* ssc install drdid, replace

* ── Load data (export from R first, or use a Stata .dta) ────────────────────
* Here we assume mpdta.dta has been written by R with:
*   haven::write_dta(mpdta, "mpdta.dta")
use "mpdta.dta", clear

xtset countyreal year
```

```{stata stata-dripw, eval=FALSE}
* ── A. Default: DR-IPW, never-treated control ────────────────────────────────
csdid lemp, ivar(countyreal) time(year) gvar(first_treat) ///
    method(dripw) notyet(0)

* Aggregate to overall ATT
estat simple
```

```{stata stata-ipw, eval=FALSE}
* ── B. IPW only ──────────────────────────────────────────────────────────────
csdid lemp, ivar(countyreal) time(year) gvar(first_treat) ///
    method(ipw) notyet(0)
estat simple
```

```{stata stata-reg, eval=FALSE}
* ── C. Regression only ───────────────────────────────────────────────────────
csdid lemp, ivar(countyreal) time(year) gvar(first_treat) ///
    method(reg) notyet(0)
estat simple
```

```{stata stata-notyettreat, eval=FALSE}
* ── D. DR-IPW, not-yet-treated control ───────────────────────────────────────
csdid lemp, ivar(countyreal) time(year) gvar(first_treat) ///
    method(dripw) notyet(1)
estat simple
```

```{stata stata-event, eval=FALSE}
* ── E. Event-study plot ──────────────────────────────────────────────────────
csdid lemp, ivar(countyreal) time(year) gvar(first_treat) method(dripw)
csdid_plot, style(rspike)
graph export "stata_eventstudy.png", replace
```

> **Note on `Statamarkdown`.**
> Add `library(Statamarkdown)` to the setup chunk and set
> `stataexe = "/path/to/stata"` in `knitr::opts_chunk$set()` (or via the
> `STATA_EXECUTABLE` environment variable). Set `eval=TRUE` on each Stata chunk
> once Stata is available.
> Alternatively, export results from Stata as `.csv` or `.dta` and read them
> into R for a unified comparison table.

---

# Summary Comparison

```{r summary-all, results='asis'}
# Combine all R results into one master table
all_results <- bind_rows(
  # Part A: algorithms
  extract_att(fit_dr,  "did · DR")    |> mutate(Dimension = "Algorithm"),
  extract_att(fit_ipw, "did · IPW")   |> mutate(Dimension = "Algorithm"),
  extract_att(fit_reg, "did · Reg")   |> mutate(Dimension = "Algorithm"),
  # Part B: base period
  extract_att(fit_never,  "did · Never")   |> mutate(Dimension = "Base period"),
  extract_att(fit_notyet, "did · NotYet")  |> mutate(Dimension = "Base period"),
  # Part C: packages
  data.frame(Method="did · att_gt",         ATT=att_did$overall.att,
             SE=att_did$overall.se,
             CI_lower=att_did$overall.att-1.96*att_did$overall.se,
             CI_upper=att_did$overall.att+1.96*att_did$overall.se,
             Dimension="Package"),
  data.frame(Method="fixest · Sun-Abraham", ATT=att_sa, SE=se_sa_agg,
             CI_lower=att_sa-1.96*se_sa_agg, CI_upper=att_sa+1.96*se_sa_agg,
             Dimension="Package"),
  data.frame(Method="staggered · R&S",      ATT=fit_stag$estimate, SE=fit_stag$se,
             CI_lower=fit_stag$estimate-1.96*fit_stag$se,
             CI_upper=fit_stag$estimate+1.96*fit_stag$se,
             Dimension="Package")
) |>
  mutate(across(where(is.numeric), ~round(., 4))) |>
  select(Dimension, Method, ATT, SE, CI_lower, CI_upper)

kable(
  all_results,
  caption = "Master Comparison: All Dimensions",
  booktabs = TRUE, linesep = ""
) |>
  kable_styling(latex_options = c("hold_position", "scale_down")) |>
  pack_rows(index = table(all_results$Dimension)) |>
  column_spec(1, bold = TRUE)
```

```{r summary-plot, fig.height=5}
ggplot(all_results,
       aes(x = reorder(Method, ATT), y = ATT,
           ymin = CI_lower, ymax = CI_upper,
           colour = Dimension)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_pointrange(size = 0.7, linewidth = 0.7) +
  coord_flip() +
  facet_wrap(~Dimension, scales = "free_y", ncol = 1) +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  labs(title = "Overall ATT: Master Comparison",
       x = NULL, y = "ATT (95% CI)") +
  theme_bw(base_size = 11)
```

---

# Session Info

```{r session}
sessionInfo()
```
